Can We Crunch the Social Graph on the Cheap?

### 2011-01-21 doc/hotpar2011/softxmt-hotpar2011.tex

\title{A latency-tolerant runtime for large graph computations}

@@

We want the best of both worlds---we want to build a system that has
good performance on both these low-locality graph codes as well as on
common scientific codes. Our approach is inspired by the XMT, but
implemented in software on commodity microprocessors. The core of our
approach is a lightweight coroutine library designed to overlap
long-latency memory prefetches with other contexts' computation. This
document describes our approach.

@@

\todo{ discussion of coroutines \& synchronization } Were a coroutine to
block on a synchronization variable shared with other threads, the
entire fray would suspend execution.  This can lead to deadlock when,
for example, one coroutine waits to consume from another thread that
is waiting to consume what only another coroutine in this first fray
can produce.  Instead, coroutines must yield on failed synchronization
events, spinning rather than blocking, where were they bonafide
thereads, blocking might be more efficient.

@@

many many threads

prefetch and switch on

full/empty bits

\subsection{Opportunities}

characteristics:

little FP use ==> don't worry about FP registers

would like enough concurrency to keep pipeline filled, but commodity CPUs don't have enough external bandwidth. So, get enough to saturate bandwidth.

@@

Describe general approach, explaining

@@

switch only stack pointer and program counter; depend on compiler to save/restore anything else. in general case, hopefully this will be fast.

\subsection{Memory concurrency}

Commodity procs don't provide enough. use software? FPGA?

@@

use existing atomic instructions? Use FPGA listening snoops on QPI?

parition remote memory on each node; have one thread access each
partition, making sync easier

\subsection{Extending to a multi-node system}

@@

current: just context switch on single node

linked lists:

there is some room for memory concurrency

thread contexts don't take up too much space in cache

sparse matmul:

speedup?

>>>>

\title{Can We Crunch the Social Graph on the Cheap?}

@@

We want the best of both worlds---we want to build a system that has
good performance on both these low-locality graph codes as well as on
common scientific codes. Our approach is inspired by the XMT, but
implemented in software on commodity microprocessors. The core of our
approach is a lightweight coroutine library designed to overlap
long-latency memory prefetches with other contexts' computation.

\todo{something like: project still in early stages. this document lays out
  problems we think a system like this needs to solve, and describes
  some implementation strategies we want to explore. we have a basic
  initial implementation with some experiments that suggest there is
  promise here.}

@@

\todo{ discussion of coroutines \& synchronization } Were a coroutine to
block on a synchronization variable shared with other threads, the
entire fray would suspend execution.  This can lead to deadlock when,
for example, one coroutine waits to consume from another thread that
is waiting to consume what only another coroutine in this first fray
can produce.  Instead, coroutines must yield on failed synchronization
events, spinning rather than blocking, where were they bona fide
threads, blocking might be more efficient.

@@

@@

Our goal is to enable high throughput on graphs with two challenging
properties. First, the graphs' combination of sparseness and low
degree means there is little locality to be exploited. Second, the
graphs are larger than a single node's memory and do not partition
well \todo{explain?}. We must spread the graph across multiple nodes
and send memory operations over a network. If memory requests are
uniformly distributed over the working set, a given node will make as
many accesses to a distant node's memory as it does to the local node.

We would like to run as fast as current hardware will let us. Pin
bandwidth is the tightest resource: Intel's QuickPath Interconnect
(QPI) allows approximately 100 million requests per second
\todo{cite}, and recent interconnect products allow similar request
rates, with a round-trip latency of approximately 3$\mu$s
\todo{cite}. Little's law tells us then that we must support on the
order of 300 concurrent memory operations, and thus coroutines, to
cover this latency.

These requirements drive the design of our system. There are three key
components.

@@

There are three main concerns in designing the coroutine
mechanism. First, contexts must be small, so many coroutines can be
active without monopolizing the cache. Second, context switches must
be fast, so that overhead does not swamp actual work. Third, when a
coroutine yields, we must choose a good coroutine to run next.

We minimize context size by treating context switches as function
calls, as in \cite{charm}. This allows us to save and restore the
minimum number of registers allowed by the ABI, and depend on the
compiler to save and restore any other registers in use. We minimize
context switch time both by keeping contexts small and by doing all
switching and scheduling in user space, as in \todo{cite other green
  threads packages}.

The choice of which coroutine to run next presents us with a
tradeoff. The simplest scheduler chooses coroutines to execute in
round-robin fashion. But this may lead to performance problems: a
coroutine might be scheduled before the event which caused it to yield
has completed, stalling the pipeline and keeping other coroutines from
progressing. A coroutine might also be scheduled long after the event
which caused it to yield has completed; while this wouldn't block
other coroutines from proceeding, it would add unnecessary latency to
the yielding coroutine. A more intelligent scheduler might allow
coroutines to execute only when they were ready, but this could
increase the overhead of context switching. Our initial implementation
takes the simple approach, but more investigation is warranted.

\subsection{Accessing Global Memory}

% Our goal is to support many concurrent accesses to a global address
% space spread across nodes in a cluster. We can draw on hardware
% support developed for the Partitioned Global Address Space family of
% languages to accomplish this.

% We draw on libraries and network support 
% goal. Languages in the Partitioned Global Address Space family such as
% X10 and Chapel have similar goals,

% It is also useful to access this low-locality data without disturbing
% the thread contexts and local storage in the cache. Current x86
% processors support non-temporal loads and stores. These memory
% operations are quite relevant for our problem: they are designed to
% move data without polluting the cache. Using these instructions works
% better than mapping the low-locality data as uncached, since that
% restricts the processors' ability to reorder operations.

Following the lead of the PGAS family of languages, we will use
the RDMA functionality provided by modern Infiniband networks to
access our global memory. \todo{rethink?}

These networks allow us to read and write memory directly from remote
nodes, but they do so through a low-level interface designed to be
hidden behind a library. Loads and stores to global memory must be
replaced with calls to a library. This library causes the network
interface to fetch the remote data and place it in a temporary
buffer. When the requesting coroutine is scheduled again, data is
returned from the buffer and execution proceeds.

In order to minimize memory access overhead, the memory operation
library calls do not interact with the network interface
directly. Instead, they queue their operations to another thread, the
{\em memory manager thread}, running on another core in the CPU. This thread
translates the global address into a network destination and issues
the network operation, checks for completion, and manages the returned
data buffers.

But we have to be careful---directing memory requests from multiple
compute threads to a single memory manager thread may reduce memory
concurrency. And we've increased the overhead of a single memory
operation---what in the single node case was a simple read, is now
multiple queuing operations between threads, and multiple reads and
writes over QPI to the network interface. Adding more memory manager
threads may enable more memory concurrency, but the overhead is still high.

Instead, we propose hardware acceleration with a {\em memory manager
  FPGA}, a coprocessor FPGA inserted into a Xeon socket so it can
participate in the node's coherence domain. This coprocessor does no
computation---it would only manage the global memory requests of the
main CPU, performing the function of the memory manager thread
described earlier.

By participating in the coherence domain, the memory manager FPGA can
act as a proxy for the rest of global memory. It advertises the
address space of the rest of the system, and translates local loads
and stores to addresses on remote nodes into the network operations
required.

Unfortunately, current processors do not provide sufficient hardware
resources for us to reach our concurrency target using this
approach. Each core in Intel's X5650 Xeon supports only 10 outstanding
misses from its private L2; the entire chip is limited to 32 L2 misses
to local memory and 12 L2 misses for memory reached over the QPI
bus. A simple translation of a local load to a network access would
tie up one of these slots for the entire multi-microsecond latency of
the request. We must bypass these structures to allow the amount of
concurrency we need; the CPU needs to direct the memory manager FPGA
to do work on its behalf without tying up its own resources.

To accomplish this, we encode commands in the high-order bits of
remote memory addresses. The memory manager FPGA advertises each
remote address twice. The first address is used for blocking reads and
writes---any access incurs the full latency of the remote
access. \todo{verify writes block?} The second address is used for
prefetches and asynchronous writes. These addresses differ by one
high-order bit, easily set with a mask in a macro. A read from the
prefetch address starts the network transaction, but immediately
returns a dummy value. When the value arrives, it is stored in a
temporary buffer; when the requesting coroutine is scheduled again, it
executes the blocking read, and the memory manager immediately returns
the value from the buffer. \todo{this isn't fully baked---do we want
  to include this sort of detail?}

@@

In systems with hundreds of threads per node, lightweight
synchronization is important. While we have not implemented any
synchronization primitives yet, full-empty bits are a natural choice
given our programming model. Two questions then arise: at what
granularity should full-empty bits work, and what component should do
the synchronization?

We expect most data accesses to be to 64-bit words. It is possible to
support full-empty bit synchronization on arbitrary words by
allocating additional storage for the full-empty bits, and translating
each atomic full-empty operation into a sequence of memory operations
that modify the data and full-empty bit atomically.

One potential optimization is to limit full-empty synchronization to
pointers to aligned 64-bit words. This leaves the bottom three bits
free to store the full-empty bits. These bits would be masked out when
the pointer was returned to the user.

These techniques can be implemented in software; \cite{qthreads} is
one example. Alternatively, atomic operations could be delegated to a
memory manager FPGA as described in the previous section. The address
command encoding would be extended to support the various full-empty
operations.

It might seem natural to use the FPGA as a memory controller as
well, storing all full-empty-bit-capable memory in DRAMs attached to
the FPGA. While this is conceptually simple, it does make the
programming model and the design more complicated. Another possibility
would be to store only the full-empty bits in memories attached to the
FPGA, and leave data in memory attached to the CPU. We believe that
starting simple is best: both data and full-empty bits should be
stored in the CPU's memory. \todo{discuss QPI bandwidth differences}

@@
